Three distinct patterns of mental response after mountain sport accidents

### 2023-03-30 paper/markdown/supplementary_material.Rmd
Changes in figures, tables and paper text
Changes in the table layout, shortening of the text, focus on early predictors of mental problems post accident

---
title: "Three distinct patterns of mental response after a sport accident"
subtitle: "Supplementary Material"
author: "Psychiatry Study Team"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

@@

Items of the EUROHIS QOL 8 tool were scores as with 1 - 5 likert scales (1: no concerns/full satisfaction, 5: extreme concerns, no satisfaction at all) with each item representing a single domain of quality of life (quality of life, health, energy, finances, activity, self-esteem, relationship and housing). 
The total EUROHIS QOL score was defined as the arithmetic mean of all items [@Schmidt2006].

@@

Traumatic events prior to the sport accident were assessed with the DIA-X tool [@Maercker2005]. 
Direct personal experience or being a witness of a traumatic event specified by the DIA-X questionnaire or by an additional yes/no item ('other traumatic events') was scored as 1. 
Prior traumatic event was assumed with at least one DIA-X item scored with 1. 
Prior sport accidents, 
flashbacks during sport activity, 
confusion during sport activity, 
subjective need for psychological support, 
presence of psychological/psychiatric support/therapy post accident 
and presence of somatic health consequences after the accident 
were surveyed as single yes/no items. 
Flashbacks frequency was assessed in the following categories: none, more than one per month and more than one per year. 
Smoking was surveyed as a single yes/no question. 
Substance abuse was investigated with the CAGE tool with $\geq$ 2 points indicative of substantial addiction [@OBrien2008]. 
Data on hospital treatment, hospitalization and surgery were extracted from electronic patient's records. 
Injury severity was assessed with the abbreviated injury scale (AIS) [@Gennarelli2006].

@@

To assess possible selection bias, demographic, socioeconomic, clinical and accident-related parameters were compared between individuals excluded due to missingness of psychometric data or denying survey participation and participants included in he analysis (__Figure 1__). 
Categorical variables were compared by $\chi^2$ test with Cramer V effect size statistic and numeric variables were compared by Mann-Whitney test with r effect size statistic (function `compare_variables()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA)). 
Potential differences between the training and test subset of the study cohort (__Figure 1__) were analyzed in an analogical way. 
Significant and near-significant (p < 0.1) differences between the included/excluded participants and the training/test subsets are presented in __Supplementary Tables S\@ref(tab:tab-incl-excl)__ and __S\@ref(tab:tab-partition)__, respectively.

@@

Observations of the training subset of the study cohort were subjected to clustering in respect to the numeric psychometric scores (__Supplementary Table S\@ref(tab:tab-mental-battery)__). 
The score values were normalized and median-centered (function `center_data()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)). 
For clustering, the PAM (partition around medoids) algorithm [@Schubert2019] with the cosine distance measure between observations [@Drost2018] was employed. 
The clustering object was constructed with the function `kcluster()` from the  [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) package.
The choice of the clustering algorithm was motivated by its good explanatory performance measured by the fraction of explained clustering variance (ratio of between-cluster sum of squares to total sum of squares, method `var()`, [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)) and superior stability in 10-fold cross-validation [@Lange2004] (cluster assignment in the folds by an inverse distance weighted 7-nearest neighbors classifier, method `cv()`, [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)) in a comparison with several other clustering algorithms presented in __Supplementary Figure S\@ref(fig:fig-clust-dev)A__. 
The number of cluster was chosen based on the bend of the curve of within-cluster sum of squares and the peak of mean silhouette statistic (method `plot()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA)). 
By this means, three mental clusters were defined: the neutral, PTG (post-traumatic growth) and PTSD (post-traumatic stress disorder) clusters (__Supplementary Figure S\@ref(fig:fig-clust-dev)B__). 
Assignment of the training subset observations to the mental clusters was accomplished with an inverse distance weighted 7-nearest neighbor classifier. 
This semi-supervised procedure yielded clustering structures with similar fractions of explained variance in the training (V = `r signif(semi_clust$variance$variance[1], 2)`) and test subset (V = `r signif(semi_clust$variance$variance[2], 2)`). 
In addition, comparable separation between the clusters could be discerned in the training and test subsets by a visual analysis of UMAP layouts, pairwise distance heat maps (__Supplementary Figure S\@ref(fig:fig-semi-clust)__) and heat maps of normalized levels of the clustering variables (__Supplementary Figure S\@ref(fig:fig-feat-hm)__).

@@

Univariable classifiers of the mental cluster assignment for demographic, socioeconomic, clinical and accident-related parameters were trained in the training subset of the study cohort by the one-rule (oneR) algorithm (function `OneR()`, package _OneR__) [@VonJouanne-Diedrich2017; @Holte1993]. 
The cluster assignment was subsequently predicted for the test subset observations (method `predict()`, package _OneR_). 
The multi-parameter classifier was developed in the training subset with the conditional random forest algorithm [@Hothorn2006; @Strobl2007; @Hothorn2022] with n = 1000 random trees, the splitting rule corresponding to the maximum test statistic and five variable draws per tree (function `cforest()` with the `cforest_control(teststat = 'max', testtype = 'Teststatistic', mtry = 5, mincriterion = qnorm(0.9), ntree = 1000)` control object, package _party_). 
The optimal variable draw values were obtained by optimization (tuning) employing the out-of-bag predictions in the training subsets (not shown). 
The cluster assignment was subsequently predicted for the test subset observations (method `predict()`, package _party_).

@@

```{r tab-mental-battery, tab.cap = 'Mental health assessment battery. The table is available in a supplementary Excel file.'}

suppl_tables$mental_battery[1, 1] %>% 
  mutate(Section = '') %>% 
  set_names(' ') %>% 
  flextable

@@

```{r tab-incl-excl, tab.cap = 'Significant and near-significant (p < 0.1) differences between individuals excluded from analysis and analyzed study participants. Numeric variables are presented as medians with interquartile ranges (IQR). Categorical variables are presented as percentages and counts within the complete observation set.'}

suppl_tables$incl_excl %>% 
  map_dfc(stri_replace, fixed = '\nrange', replacement = ', range') %>% 
  flextable %>% 
  flextable::width(1, width = 3.2, unit = 'cm') %>% 
  flextable::width(2:3, width = 5.5, unit = 'cm') %>% 
  flextable::width(4, width = 2.7, unit = 'cm') %>% 
  flextable::width(5, width = 2.3, unit = 'cm') %>% 
  footnote(1, 4:5, 
           value = as_paragraph('Numeric variables: Mann-Whitney test with r effect size statistic; categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(8, 1, 
           value = as_paragraph('AIS: abbreviated injury scale'), 
           part = 'body', 
           ref_symbols = 'b') %>% 
  footnote(11, 1, 
           value = as_paragraph('PCL-5 DSM-5: PTSD checklist for DSM-5'), 
           part = 'body', 
           ref_symbols = 'c') %>% 
  footnote(12:15, 1, 
           value = as_paragraph('PTGI: post-traumatic growth inventory'), 
           part = 'body', 
           ref_symbols = 'd') %>% 
  footnote(16, 1, 
           value = as_paragraph('PSS4: 4-item perceived stress scale'), 
           part = 'body', 
           ref_symbols = 'e') %>% 
  footnote(17, 1, 
           value = as_paragraph('SOC-9L: Leipzig 9-item sense of coherence scale'), 
           part = 'body', 
           ref_symbols = 'f') %>% 
  theme_vanilla

```{r tab-partition, tab.cap = 'Significant and near-significant (p < 0.1) differences between the training and test subsets of the study cohort. Numeric variables are presented as medians with interquartile ranges (IQR). Categorical variables are presented as percentages and counts within the complete observation set.'}

flextable::flextable(suppl_tables$partition) %>% 
  flextable::width(1, width = 3.2, unit = 'cm') %>% 
  flextable::width(2:3, width = 5.5, unit = 'cm') %>% 
  flextable::width(4, width = 2.7, unit = 'cm') %>% 
  flextable::width(5, width = 2.3, unit = 'cm') %>% 
  footnote(1, 4:5, 
           value = as_paragraph('\u03C7\u00B2 test with Cramer V effect size statistic'), 
           part = 'body', 
           ref_symbols = 'a') %>% 
  footnote(2, 1, 
           value = as_paragraph('RS13: 13-item resilience scale'), 
           part = 'body', 
           ref_symbols = 'b') %>% 
  theme_vanilla

@@

```{r tab-clust-bcg, tab.cap = 'Significant and near-significant (p < 0.1) differences in demographic, socioeconomic, clinical and accident-related factors between the mental clusters. Numeric variables are presented as medians with interquartile ranges (IQR). Categorical variables are presented as percentages and counts within the complete observation set. The table is available in a supplementary Excel file.'}

suppl_tables$clust_bcg[1, 1] %>% 
  mutate(Variable = '') %>% 
  set_names(' ') %>% 
  flextable

@@

_For categorical variables, percentages of complete observations are presented as bars. Numeric variables are presented in violin plots with red diamonds denoting medians, interquartile ranges presented as red whiskers and single observations depicted as points. The number of complete cases is displayed in the plot captions._ 
_(A) Distribution of injured body regions. (B) Scoring of the domains of the EUROHIS project 8-item quality of life scale (EUROHIS-QOL 8). (C) Scoring of the post-traumatic stress disorder (PTSD) clusters with the PCL-5 DSM-5 tool and percentages of participants screened positive for the PTSD clusters. (E) Scoring of the factors of the post-traumatic growth with the post-traumatic growth inventory (PTGI) tool._

@@

_The mental clusters were defined in respect to psychometric scoring in the training subset of the study cohort by PAM (partition around medoids) with cosine distance between the observations. Assignment of the test subset observations to the mental clusters was done with the inverse distance weighted 7-nearest neighbors classifier. By this procedure, three mental clusters were identified: neutral, PTG (post-traumatic growth) and PTSD (post-traumatic stress disorder). Numbers of observations in the mental clusters are shown in the plot legend in (A)._

@@

```{r fig-oneR, fig.width = mm_inch(suppl_figures$oneR$w), fig.height = mm_inch(suppl_figures$oneR$h), fig.cap = 'Prediction of the mental cluster assignment by single demographic, socioeconomic, clinical and accident-related factors.'}

@@

__Supplementary Figure S\@ref(fig:fig-oneR). Prediction of the mental cluster assignment by single demographic, socioeconomic, clinical and accident-related factors.__

_Univariable classifiers of the mental cluster assignment for demographic, socioeconomic, clinical and accident-related parameters were trained in the training subset of the study cohort by the one-rule (oneR) algorithm. The cluster assignment was subsequently predicted for the test subset observations. Accuracy and $\kappa$ of the classifiers with any predictive value ($\kappa$ > 0 in both subsets) are shown in bar plots._

@@

```{r fig-early-crf, fig.width = mm_inch(suppl_figures$early_crf$w), fig.height = mm_inch(suppl_figures$early_crf$h), fig.cap = 'Assignment of accident victims to the mental clusters based on explanatory factors available during acute medical management of the accident.'}

suppl_figures$early_crf$plot

@@

__Supplementary Figure S\@ref(fig:fig-early-crf). Assignment of accident victims to the mental clusters based on explanatory factors available during acute medical management of the accident.__

_A conditional random forest classifier for the mental cluster assignment based on demographic, socioeconomic, clinical and accident-related factors available during early acute medical management of the accident was trained in the training subset of the study cohort._

>>>>

---
title: "Three distinct patterns of mental response after mountain sport accidents"
subtitle: "Supplementary Material"
author: "Psychiatry Study Team"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

@@

Items of the EUROHIS QOL 8 tool were scores as with 1 - 5 likert scales (1: extreme concerns, no satisfaction at all, 5: no concerns/full satisfaction, ) with each item representing a single domain of quality of life (quality of life, health, energy, finances, activity, self-esteem, relationship and housing). 
The total EUROHIS QOL score was defined as the arithmetic mean of all items [@Schmidt2006].

@@

Traumatic events prior to the sport accident were assessed with the DIA-X tool (Diagnostic Expert System) [@Maercker2005]. 
Direct personal experience or being a witness of a traumatic event specified by the DIA-X questionnaire or by an additional yes/no item ('other traumatic events') was scored as 1. 
Prior traumatic event was assumed with at least one DIA-X item scored with 1. 
Prior sport accidents, 
flashbacks during sport activity, 
confusion during sport activity, 
subjective need for psychological support, 
presence of psychological/psychiatric support/therapy post accident 
and presence of somatic health consequences after the accident 
were surveyed as single yes/no items. 
Flashbacks frequency was assessed in the following categories: none, more than one per month and more than one per year. 
Smoking was surveyed as a single yes/no question. 
Alcohol abuse was investigated with the CAGE tool with $\geq$ 2 points indicative of substantial addiction [@OBrien2008]. 
Data on hospital treatment and surgery were extracted from electronic patient's records. 
Injury severity was assessed with the abbreviated injury scale (AIS) [@Gennarelli2006].

@@

To assess possible selection bias, demographic, socioeconomic, clinical and accident-related parameters were compared between individuals excluded due to missingness of psychometric data or denying survey participation and participants included in he analysis (__Figure 1__). 
Categorical variables were compared by $\chi^2$ test with Cramer V effect size statistic and numeric variables were compared by Mann-Whitney test with r effect size statistic (function `compare_variables()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA)). 
Potential differences between the training and test subset of the study cohort (__Figure 1__) were analyzed in an analogical way. 
Significant (p < 0.05) differences between the included/excluded participants are presented in __Supplementary Tables S\@ref(tab:tab-incl-excl)__.

@@

Observations of the training subset of the study cohort were subjected to clustering in respect to the numeric psychometric scores (__Supplementary Table S\@ref(tab:tab-mental-battery)__). 
The score values were normalized and median-centered (function `center_data()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)). 
For clustering, the PAM (partition around medoids) algorithm [@Schubert2019] with the cosine distance measure between observations [@Drost2018] was employed. 
The clustering object was constructed with the function `kcluster()` from the  [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) package.
The choice of the clustering algorithm was motivated by its good explanatory performance measured by the fraction of explained clustering variance (ratio of between-cluster sum of squares to total sum of squares, method `var()`, [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)) and superior stability in 10-fold cross-validation [@Lange2004] (cluster assignment in the folds by an inverse distance weighted 7-nearest neighbors classifier, method `cv()`, [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)) in a comparison with several other clustering algorithms presented in __Supplementary Figure S\@ref(fig:fig-clust-dev)A__. 
The number of cluster was chosen based on the bend of the curve of within-cluster sum of squares and the peak of mean silhouette statistic (method `plot()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA)). 
By this means, three mental clusters were defined: the neutral, PTG (post-traumatic growth) and PTB (post-traumatic burden) clusters (__Supplementary Figure S\@ref(fig:fig-clust-dev)B__). 
Assignment of the training subset observations to the mental clusters was accomplished with an inverse distance weighted 7-nearest neighbor classifier. 
This semi-supervised procedure yielded clustering structures with similar fractions of explained variance in the training (V = `r signif(semi_clust$variance$variance[1], 2)`) and test subset (V = `r signif(semi_clust$variance$variance[2], 2)`). 
In addition, comparable separation between the clusters could be discerned in the training and test subsets by a visual analysis of UMAP layouts, pairwise distance heat maps (__Supplementary Figure S\@ref(fig:fig-semi-clust)__) and heat maps of normalized levels of the clustering variables (__Supplementary Figure S\@ref(fig:fig-feat-hm)__).

@@

Univariable classifiers of the mental cluster assignment for demographic, socioeconomic, clinical and accident-related parameters (__Supplementary Table S\@ref(tab:tab-mod-variables)__) were trained in the training subset of the study cohort by the one-rule (oneR) algorithm (function `OneR()`, package _OneR__) [@VonJouanne-Diedrich2017; @Holte1993]. 
The cluster assignment was subsequently predicted for the test subset observations (method `predict()`, package _OneR_).

Two multi-parameter classifiers of the mental cluster assignment were developed in the training subset:

1. a model employing candidate early predictors of cluster assignment, i.e. demographic, socioeconomic, medical history and accident related explanatory factors

2. a model including additionally late consequences of the accident such as somatic health aftermath, flashbacks or cautious behavior during sport

The multi-parameter classifiers (__Supplementary Table S\@ref(tab:tab-mod-variables)__) were trained in the training subset with the conditional random forest algorithm [@Hothorn2006; @Strobl2007; @Hothorn2022] with n = 1000 random trees and the splitting rule corresponding to the maximum test statistic. 
In the model 1, the `mtry` parameter was set to 7, `minsplit` to 10 and `mincriterion` to 1.28 (function `cforest()` with the `cforest_control(teststat = 'max', testtype = 'Teststatistic', mtry = 7, mincriterion = qnorm(0.9), minsplit = 10, ntree = 1000)` control object, package _party_). 
In the model 2, the `mtry` parameter was set to 6, `minsplit` to 20 and `mincriterion` to 0 (function `cforest()` with the `cforest_control(teststat = 'max', testtype = 'Teststatistic', mtry = 6, mincriterion = 0, minsplit = 20, ntree = 1000)` control object, package _party_). 
The optimal parameter sets were obtained by optimization (tuning) employing the out-of-bag predictions in the training subsets (not shown). 
The cluster assignment was subsequently predicted for the test subset observations (method `predict()`, package _party_).

@@

```{r tab-mental-battery, tab.cap = 'Mental health assessment battery.'}

suppl_tables$mental_battery[c(2, 3)] %>% 
  flextable %>% 
  width(1, width = 6.5, unit = 'cm') %>% 
  width(2, width = 10.5, unit = 'cm') %>% 
  theme_vanilla

@@

```{r tab-incl-excl, tab.cap = 'Significant (p < 0.05) differences between individuals excluded from analysis and analyzed study participants. Numeric variables are presented as medians with interquartile ranges (IQR). Categorical variables are presented as percentages and counts within the complete observation set.'}

suppl_tables$incl_excl %>% 
  map_dfc(stri_replace, fixed = '\nrange', replacement = ', range') %>% 
  flextable %>% 
  flextable::width(1, width = 3.2, unit = 'cm') %>% 
  flextable::width(2:3, width = 5.5, unit = 'cm') %>% 
  flextable::width(4, width = 2.7, unit = 'cm') %>% 
  flextable::width(5, width = 2.3, unit = 'cm') %>% 
  footnote(1, 1, 
           value = as_paragraph('AIS: abbreviated injury scale; PTGI: post-traumatic growth inventory; SOC-9L: Leipzig 9-item sense of coherence scale; K: 1000 Euro; EUR: Euro.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 4:5, 
           value = as_paragraph('Numeric variables: Mann-Whitney test with r effect size statistic; categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic, '), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  theme_vanilla

@@

```{r tab-clust-bcg, tab.cap = 'Significant (p < 0.05) differences in demographic, socioeconomic, clinical and accident-related factors between the mental clusters. Numeric variables are presented as medians with interquartile ranges (IQR). Categorical variables are presented as percentages and counts within the complete observation set.'}

suppl_tables$clust_bcg %>% 
  flextable %>% 
  width(1, width = 1.7, unit = 'cm') %>% 
  width(2, width = 3.7, unit = 'cm') %>% 
  width(3:5, width = 3.5, unit = 'cm') %>% 
  width(6, width = 2.4, unit = 'cm') %>% 
  width(7, width = 2.2, unit = 'cm') %>% 
  merge_v(1) %>% 
  footnote(1, 2, 
           value = as_paragraph('CAGE: Cut, Annoyed, Guilty, and Eye scale; AIS: abbreviated injury score.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 6:7, 
           value = as_paragraph('Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  theme_vanilla

```{r tab-mod-variables, tab.cap = 'Sets of explanatory factors (early: available during acute medical management of the accident) used for modeling of the mental cluster assignment.'}

flextable::flextable(suppl_tables$mod_variables) %>% 
  width(1, width = 3, unit = 'cm') %>% 
  width(2, width = 12, unit = 'cm') %>% 
  theme_vanilla

@@

_For categorical variables, percentages of complete observations are presented as bars. Numeric variables are presented in violin plots with red diamonds denoting medians, interquartile ranges presented as red whiskers and single observations depicted as points. Quality of life scores are expressed as categorical variables and presented as percentages of participants in stack plots. The number of complete cases is displayed in the plot captions._ 
_(A) Distribution of injured body regions. (B) Scoring of the domains of the EUROHIS project 8-item quality of life scale (EUROHIS-QOL 8). (C) Scoring of the post-traumatic stress disorder (PTSD) domains with the PCL-5 DSM-5 tool and percentages of participants screened positive for the PTSD domains. (E) Scoring of the factors of the post-traumatic growth with the post-traumatic growth inventory (PTGI) tool._

@@

_The mental clusters were defined in respect to psychometric scoring in the training subset of the study cohort by PAM (partition around medoids) with cosine distance between the observations. Assignment of the test subset observations to the mental clusters was done with the inverse distance weighted 7-nearest neighbors classifier. By this procedure, three mental clusters were identified: neutral, PTG (post-traumatic growth) and PTB (post-traumatic burden). Numbers of observations in the mental clusters are shown in the plot legend in (A)._

@@

```{r fig-oneR, fig.width = mm_inch(suppl_figures$oneR$w), fig.height = mm_inch(suppl_figures$oneR$h), fig.cap = 'Prediction of the mental cluster assignment by single demographic, socioeconomic, clinical and accident-related factors available during acute medical management of the accident.'}

@@

__Supplementary Figure S\@ref(fig:fig-oneR). Prediction of the mental cluster assignment by single demographic, socioeconomic, clinical and accident-related factors available during acute medical management of the accident.__

_Univariable classifiers of the mental cluster assignment for demographic, socioeconomic, clinical and accident-related parameters available during acute medical management of the mountain sport accident were trained in the training subset of the study cohort by the one-rule (oneR) algorithm. The cluster assignment was subsequently predicted for the test subset observations. Accuracy and $\kappa$ of the classifiers with any predictive value ($\kappa$ > 0 in both subsets) are shown in bar plots._

@@

```{r fig-crf, fig.width = mm_inch(suppl_figures$crf$w), fig.height = mm_inch(suppl_figures$crf$h), fig.cap = 'Assignment of accident victims to the mental clusters based on demographic, socioeconomic, clinical, accident- and accident consequence-related factors by the random forest algorithm.'}

suppl_figures$crf$plot

@@

__Supplementary Figure S\@ref(fig:fig-crf). Assignment of accident victims to the mental clusters based on demographic, socioeconomic, clinical, accident- and accident consequence-related factors by the random forest algorithm.__

_A conditional random forest classifier for the mental cluster assignment based on demographic, socioeconomic, clinical and accident- and accident consequence-related factors was trained in the training subset of the study cohort._
